``Interestingness'' in Mathematical Theory Exploration

### 2019-06-02 PhDThesis/appendix_nixeval.tex
Thesis updates

\begin{sloppypar}
  Haskell's support for runtime evaluation of code (\'a la Lisp's
  \eval{}~\cite[\S~1.6]{edwards1962lisp}) is limited. The most popular mechanism
  is the \hint{} package\footnote{https://hackage.haskell.org/package/hint},
  which is built on top of GHC's API and provides functions like
  \hs{eval :: MonadInterpreter m => String -> m String}. The
  \hs{MonadInterpreter} constraint provides mechanisms for calling into GHC and
  managing its ``session'', which includes details like compiler flags, a list
  of imported modules and paths to compiled packages.

@@

  To solve this problem we created the \nixeval{}
  package\footnote{https://hackage.haskell.org/package/nix-eval}. This provides
  a function \hs{eval :: Expr -> IO (Maybe String)}, which produces \hs{String}s
  for the same reasons as \hint{}, but differs in its handling of expressions
  and context. Rather than operating on a raw \hs{String}, \nixeval{} defines
  the \hs{Expr} type, containing several fields to keep track of the context in
  which the expression is to be evaluated:
\end{sloppypar}

@@

-- Forces a value's type, serialises it then hashes it
hasher :: Expr
hasher = compose [withPkgs ["murmur-hash"]
                    (qualified "Data.Digest.Murmur32" "asWord32"),
                  withPkgs ["murmur-hash"]
                    (qualified "Data.Digest.Murmur32" "hash32"),
                  withPkgs ["cereal"]
                    (qualified "Data.Serialize" "runPut"),
                  withPkgs ["cereal"]
                    (qualified "Data.Serialize" "put"),
                  converter]
  \end{haskell}
  \caption{Example usage of \nixeval{} functions, taken from \textsc{ML4HS}.
    These are used to dynamically generate \qspec{} signatures of user-provided
    functions.}
  \label{fig:nix-eval}
\end{figure}

>>>>

\begin{sloppypar}
  Haskell's support for runtime evaluation of code (\'a la Lisp's
  \eval{}~\cite[\S~1.6]{edwards1962lisp}) is limited. The most popular mechanism
  is the \hint{} package\footnote{\hPackage{hint}}, which is built on top of
  GHC's API and provides functions like
  \hs{eval :: MonadInterpreter m => String -> m String}. The
  \hs{MonadInterpreter} constraint provides mechanisms for calling into GHC and
  managing its ``session'', which includes details like compiler flags, a list
  of imported modules and paths to compiled packages.

@@

  To solve this problem we created the \nixeval{}
  package\footnote{\hPackage{nix-eval}}. This provides a function
  \hs{eval :: Expr -> IO (Maybe String)}, which produces \hs{String}s for the
  same reasons as \hint{}, but differs in its handling of expressions and
  context. Rather than operating on a raw \hs{String}, \nixeval{} defines the
  \hs{Expr} type, containing several fields to keep track of the context in
  which the expression is to be evaluated:
\end{sloppypar}

@@

-- Forces a value's type, serialises it then hashes it
hasher :: Expr
hasher = compose [withPkgs ["murmur-hash"]
                    (qualified "Data.Digest.Murmur32" "asWord32"),
                  withPkgs ["murmur-hash"]
                    (qualified "Data.Digest.Murmur32" "hash32"),
                  withPkgs ["cereal"]
                    (qualified "Data.Serialize" "runPut"),
                  withPkgs ["cereal"]
                    (qualified "Data.Serialize" "put"),
                  converter]
  \end{haskell}
  \caption{Example usage of \nixeval{} functions, taken from \mlforhs{}.
    These are used to dynamically generate \quickspec{} signatures of
    user-provided functions.}
  \label{fig:nix-eval}
\end{figure}

### 2019-06-02 PhDThesis/background_conjecturegeneration.tex
Headingy things
Thesis updates

\subsection{Conjecture Generation}

\begin{figure}
  \begin{equation*}
    \begin{split}
      \forall a. \Nil            &: \ListA                                  \\
      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
      \Head(\Cons(x, xs))        &= x                                       \\
      \Tail(\Cons(x, xs))        &= xs                                      \\
      \Append(\Nil,         ys)  &= ys                                      \\
      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
      \Reverse(\Nil)             &= \Nil                                    \\
      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
      \Length(\Nil)              &= \Zero                                   \\
      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
      \Map(f, \Nil)              &= \Nil                                    \\
      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
      \Foldl(f, x, \Nil)         &= x                                       \\
      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
      \Foldr(f, \Nil,         y) &= y                                       \\
      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\List$ type and some associated
    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
  \label{figure:list_theory}
\end{figure}

We focus on the task of generating conjectures from some given mathematical
theory. This theory may be, for example, a scientific model, an executable
computer program or simply an object of mathematical curiosity. A theory
defines (perhaps implicitly) a particular system of logic in which to work (for
example higher-order logic), along with a finite set of \emph{definitions} (like
those in figure~\ref{figure:list_theory}) which assign structure or meaning to
particular terms. Theoretically, this is all of the information we need to
derive the truth or falsity of all decidable statements involving these terms,
for example by enumerating all valid proofs. However, due to the exponential
search spaces involved, such brute-force enumeration is infeasible in practice;
plus the resulting proofs would be difficult to comprehend and the processed
statements would be overwhelming in number and mostly uninteresting in content.

@@

As an example, we use the theory of lists shown in
figure~\ref{figure:list_theory}. These definitions are so widely used in
software that they appear in the standard libraries of many programming
languages. An example of a conjecture involving these definitions is the
following universally quantified equation:

>>>>

\section{Conjecture Generation}

\begin{figure}
  \begin{equation*}
    \begin{split}
      \forall a. \Nil            &: \ListA                                  \\
      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
      \Head(\Cons(x, xs))        &= x                                       \\
      \Tail(\Cons(x, xs))        &= xs                                      \\
      \Append(\Nil,         ys)  &= ys                                      \\
      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
      \Reverse(\Nil)             &= \Nil                                    \\
      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
      \Length(\Nil)              &= \Zero                                   \\
      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
      \Map(f, \Nil)              &= \Nil                                    \\
      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
      \Foldl(f, x, \Nil)         &= x                                       \\
      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
      \Foldr(f, \Nil,         y) &= y                                       \\
      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\List$ type and some associated
    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
  \label{fig:list_theory}
\end{figure}

We focus on the task of generating conjectures from some given mathematical
theory. This theory may be, for example, a scientific model, an executable
computer program or simply an object of mathematical curiosity. A theory
defines (perhaps implicitly) a particular system of logic in which to work (for
example higher-order logic), along with a finite set of \emph{definitions} (like
those in figure~\ref{fig:list_theory}) which assign structure or meaning to
particular terms. Theoretically, this is all of the information we need to
derive the truth or falsity of all decidable statements involving these terms,
for example by enumerating all valid proofs. However, due to the exponential
search spaces involved, such brute-force enumeration is infeasible in practice;
plus the resulting proofs would be difficult to comprehend and the processed
statements would be overwhelming in number and mostly uninteresting in content.

@@

As an example, we use the theory of lists shown in
figure~\ref{fig:list_theory}. These definitions are so widely used in
software that they appear in the standard libraries of many programming
languages. An example of a conjecture involving these definitions is the
following universally quantified equation:

### 2019-06-02 PhDThesis/background_haskell.tex
Slight fiddling
Thesis updates

\begin{itemize}
\item The large amount of Haskell code which is freely available online, e.g. in
  repositories like \href{http://hackage.haskell.org}{Hackage}, with which we
  can experiment.

\item The existence of theory exploration systems such as \hspec{}, \qspec{} and
  \textsc{Speculate}.

\item Related tooling we can re-use such as counterexample finders (\qcheck{},
  \textsc{SmallCheck}, \textsc{SmartCheck}, \textsc{LeanCheck},
  \textsc{Hedgehog}, etc.), theorem provers
  (e.g. \textsc{Hip}~\cite{rosen2012proving}), and other testing and
  term-generating systems like \textsc{MuCheck}~\cite{le2014mucheck},
  \textsc{MagicHaskeller}~\cite{katayama2011magichaskeller} and
  \textsc{Djinn}~\cite{augustsson2005djinn}.

@@

Further evidence of Haskell's suitability for theory exploration is given by the
fact that the state-of-the-art implementation for Isabelle/HOL, the
\textsc{Hipster}~\cite{Hipster} system, is actually implemented by translating
to Haskell and invoking \hspec{}~\cite{claessen2013automating}.

>>>>

\begin{itemize}
\item The large amount of Haskell code which is freely available online, e.g. in
  repositories like \hackage{}, with which we
  can experiment.

\item The existence of theory exploration systems such as \hipspec{},
  \quickspec{} and \speculate{}.

\item Related tooling we can re-use such as counterexample finders (\quickcheck{},
  \smallcheck{}, \smartcheck{}, \leancheck{},
  \hedgehog{}, etc.), theorem provers
  (e.g. \hip{}~\cite{rosen2012proving}), and other testing and
  term-generating systems like \mucheck{}~\cite{le2014mucheck},
  \magichaskeller{}~\cite{katayama2011magichaskeller} and
  \djinn{}~\cite{augustsson2005djinn}.

@@

Further evidence of Haskell's suitability for theory exploration is given by the
fact that the state-of-the-art implementation for Isabelle/HOL, the
\hipster{}~\cite{Hipster} system, is actually implemented by translating
to Haskell and invoking \hipspec{}~\cite{claessen2013automating}.

\subsection{GHC Core}
\label{sec:ghccore}

### 2019-06-02 PhDThesis/background_quickcheck.tex
Thesis updates

Although unit testing is the de facto industry standard for quality assurance in
non-critical systems, the level of confidence it provides is rather low, and
totally inadequate for many (e.g. life-) critical systems. To see why, consider
the following Haskell function, along with some unit
tests\footnote{Backticks allow functions to be written as infix operators, so
  \hs{x `div` y} is equivalent to \hs{div x y}.}:

\begin{haskell}
factorial 0 = 1
factorial n = n * factorial (n-1)

fact_base      = factorial 0 == factorial 1
fact_increases = factorial 3 <= factorial 4
fact_div       = factorial 4 == factorial 5 `div` 5
\end{haskell}

@@

All of Haskell's built-in numeric types allow negative numbers, which this
implementation doesn't take into account. Whilst this is a rather trivial
example, it highlights a common problem: unit tests are insufficient to expose
incorrect assumptions. In this case, our assumption that numbers are positive
has caused a bug in the implementation \emph{and} limited the tests we've
written.

If we do manage to spot this error, we might capture it in a \emph{regression
  test} and update the definition of \hs{factorial} to handle negative numbers,
e.g. by taking their absolute value:

\begin{haskell}
factorial 0 = 1
factorial n = let nPos = abs n
               in nPos * factorial (nPos - 1)

fact_neg = factorial 1 == factorial (-1)
\end{haskell}

However, this is \emph{still} not enough, since this function will also accept
fractional values\footnote{Since we only use generic numeric operations, the
  function will be polymorphic with a type of the form \hs{forall t. Num t => t
    -> t}, where \hs{Num t} constrains the type variable \hs{t} to be
  numeric. In fact, Haskell will infer extra constraints such as \hs{Eq t} since
  we have used \hs{==} in the unit tests.}, which will also cause it to
diverge. Clearly, by choosing what to test we are biasing the test suite towards
those cases we've already taken into account, whilst neglecting the problems we
did not expect.

Haskell offers a partial solution to this problem in the form of \emph{property
  checking}. Tools such as \qcheck{} separate tests into three components: a
\emph{property} to check, which unlike a unit test may contain \emph{free
  variables}; a source of values to instantiate these free variables; and a
stopping criterion.

\subsection{\qcheck{}}
\label{sec:quickcheck}

\qcheck{}~\cite{quickcheck} is the most widely used property checking library
for
Haskell~\footnote{According to downloads counted at
  \href{http://hackage.haskell.org/packages/browse}, and packages depending on
  it at \href{https://packdeps.haskellers.com/reverse/QuickCheck}, accessed on
  2019-05-21.}. Here is how we might restate our unit tests as \qcheck{}
properties:

\begin{haskell}
fact_base        = factorial 0 == factorial 1
fact_increases n = factorial n <= factorial (n+1)
fact_div       n = factorial n == factorial (n+1) `div` (n+1)
fact_neg       n = factorial n == factorial (-n)
\end{haskell}

The free variables (all called \hs{n} in this case) are abstracted as function
parameters; these parameters are implicitly \emph{universally quantified},
i.e. we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a
property asserting $\forall n, factorial(n) \leq factorial(n+1)$. Notice that
unit tests like \hs{fact_base} are valid properties; they just assert rather
weak statements.

To check these properties, \qcheck{} treats closed terms (like \hs{fact_base})
just like unit tests: pass if they evaluate to \hs{True}, fail otherwise. For
open terms, a random selection of values are generated and passed in via the
function parameter; the results are then treated in the same way as closed
terms. The default stopping criterion for \qcheck{} (for each test) is when a
single generated test fails, or when 100 generated tests pass.

@@

\begin{itemize}
\item \qcheck{} cannot test polymorphic functions; they must be
  \emph{monomorphised} first (instantiated to a particular concrete type). This
  is a technical limitation, since \qcheck{} must know which type of values to
  generate, but in our example it would bring the issue with fractional values
  to our attention.

\item The generators used by \qcheck{} depend only on the \emph{type} of value
  they are generating: since \hs{Int} includes positive and negative values, the
  \hs{Int} generator will output both. This will expose the problem with
  negative numbers, which we weren't expecting.
\end{itemize}

Whilst property checking generalises and improves upon unit testing, the problem
of tests being biased towards expected cases remains, since the properties to be
checked are still manually specified. Property checking can be complemented by
\emph{theory exploration} (TE), which avoids this bias by \emph{discovering}
such properties; through a combination of brute-force enumeration, random
testing and (in the case of \hspec{} and \hipster{}) automated theorem proving.
These TE systems make use of property checkers as part of their search process,
to try and disprove generated conjectures before output (to the user, or a
subsequent tool).

>>>>

Although unit testing is the de facto industry standard for quality assurance in
non-critical systems, the level of confidence it provides is rather low, and
totally inadequate for many (e.g. life-) critical systems. To see why, consider
the following Haskell function, along with some unit
tests\footnote{Haskell functions can be prefix or infix: prefix functions have
  alphanumeric names, like \hs{div} and \hs{factorial}, and appear in function
  calls before their arguments, e.g. \hs{div 10 5}; infix functions have
  non-alphanumeric names, like \hs{+} and \hs{==}, and appear in function calls
  between their first two arguments, e.g. \hs{4 + 3}. Prefix functions can
  appear infix using backticks, like \hs{10 `div` 5}, and infix functions can
  appear prefix using parentheses, like \hs{(+) 4 3}.}:

% BEGIN unit
\begin{haskell}
factorial 0 = 1
factorial n = n * factorial (n - 1)

fact_base      = factorial 0 == factorial 1
fact_increases = factorial 3 <= factorial 4
fact_div       = factorial 4 == factorial 5 `div` 5
\end{haskell}
% END unit

@@

Haskell's built-in numeric types allow negative numbers, which this
implementation doesn't take into account. Whilst this is a rather trivial
example, it highlights a common problem: unit tests are insufficient to expose
incorrect assumptions. In this case, our assumption that numbers are positive
has caused a bug in the implementation \emph{and} limited the tests we've
written.

One way to avoid this problem, if we do manage to spot it, would make results
symmetric around zero, which can be checked by a \emph{regression test} such as:

% BEGIN neg
\begin{haskell}
fact_neg = factorial 1 == factorial (-1)
\end{haskell}
% END neg

The implementation can be modified by composing if with the \hs{abs}olute value
function:

% BEGIN abs
\begin{haskell}
factorial = original . abs
  where original 0 = 1
        original n = n * naive (n - 1)
\end{haskell}
% END abs

However, this is \emph{still} not enough, since this function will also accept
fractional values\footnote{Since we only use generic numeric operations, the
  function will be polymorphic with a type of the form
  \hs{forall t. Num t => t -> t}, where \hs{Num t} constrains the type variable
  \hs{t} to be numeric.}, which will also cause it to diverge. This problem can
be fixed by restricting \hs{factorial}'s type with an explicit annotation, e.g.
to only whole numbers: \hs{factorial :: forall t. Integral t => t -> t}

Clearly, by choosing which scenarios to check, we are biasing the test suite
towards those cases we've already thought of, whilst neglecting the problems we
did not expect. Haskell offers a partial solution to this problem in the form of \emph{property checking}. Tools such as \quickcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain
\emph{free variables}; a source of values to instantiate these free variables; and a
stopping criterion.

\subsection{\quickcheck{}}
\label{sec:quickcheck}

\quickcheck{}~\cite{claessen2011quickcheck} is the most widely used property
checking library for
Haskell~\footnote{According to downloads counted at
  \url{http://hackage.haskell.org/packages/browse}, and packages depending on
  it at \url{https://packdeps.haskellers.com/reverse/QuickCheck}, accessed on
  2019-05-21.}. Here is how we might restate our unit tests as \quickcheck{}
properties:

% BEGIN quickchecked
\begin{haskell}
fact_base        = factorial 0 == factorial 1
fact_increases n = factorial n <= factorial (n + 1)
fact_div       n = factorial n == factorial (n + 1) `div` (n + 1)
fact_neg       n = factorial n == factorial (-n)
\end{haskell}
% END quickchecked

The free variables (all called \hs{n} in this case) are abstracted as function
parameters; these parameters are implicitly \emph{universally quantified},
i.e. we've gone from a unit test asserting $factorial(3) \leq factorial(4)$ to a
property asserting $\forall n, factorial(n) \leq factorial(n+1)$. Notice that
unit tests like \hs{fact_base} are valid properties; they just happen to assert
rather weak statements, since they contain no free variables.

To check these properties, \quickcheck{} treats closed terms (like \hs{fact_base})
just like unit tests: pass if they evaluate to \hs{True}, fail otherwise. For
open terms, a random selection of values are generated and passed in via the
function parameter; the results are then treated in the same way as closed
terms. The default stopping criterion for \quickcheck{} (for each test) is when a
single generated test fails, or when 100 generated tests pass.

@@

\begin{itemize}
\item We are forced to confront the generality of \hs{factorial}'s type, since
  \quickcheck{} cannot test polymorphic functions directly; they must be
  \emph{monomorphised} first (instantiated to a particular concrete type), to
  decide which data generator to use (i.e. which instance of \quickcheck{}'s
  \hs{Arbitrary} type class). This can be done either by specifying a generator
  directly (avoiding the type class mechanism), or by adding explicit type
  annotations, or by calling \quickcheck{}'s \hs{monomorphic}
  function.\footnote{\hs{monomorphic} picks \hs{Integer} for all polymorphic
    type variables. This na\"ive choice instantiates many common type classes
    like \hs{Num}, \hs{Show} and \hs{Eq}; variables whose constraints are not
    satisfied will fail to type check (e.g. \hs{Bounded}, since \hs{Integer} is
    a bignum).} This will bring the problem of fractional values. Instead of
  monomorphising the properties, we might choose to monomorphise \hs{factorial}

\item By default, \quickcheck{} picks a generator based on the \emph{type} of
  value to be generated: since \hs{Int} includes positive and negative values,
  the \hs{Int} generator will output both. This will expose the problem with
  negative numbers, which we weren't expecting.
\end{itemize}

Whilst property checking generalises and improves upon unit testing, the problem
of tests being biased towards expected cases remains, since the properties to be
checked are still manually specified. Property checking can be complemented by
\emph{theory exploration} (TE), which avoids this bias by \emph{discovering}
such properties; through a combination of brute-force enumeration, random
testing and (in the case of \hipspec{} and \hipster{}) automated theorem
proving. Property checkers like \quickcheck{} are an important component of TE
systems, since their data generators can the search process, and they can
prevent ``obvious'' falsehoods from being output by checking them first.
